---
title: "Devoir1"
output: pdf_document
date: "2024-09-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Librairies and date importation
```{r include=FALSE}
# Load necessary packages
library(dplyr)
library(lubridate)
#library("ggplot")
```

```{r}
mydata <- read.csv("C:/Users/Admin/OneDrive/Desktop/Automne 2024/Statistical modeling/GitPractice/Practice1/MATH60604A-project-bixi_part1_team5.csv")
#head(mydata)
```
## Cleaning and data preparation
### Cleaning function
```{r}
preproc_dat <- function(mydata){
  dat <- mydata
  
  #Changing variables (mem,wday,rushhours) to factors, creating var to code for weekend and rushmod (modified rush hour that excludes weekend)
  dat <- mutate(dat, wend = as.factor(as.integer(dat$wday %in% c("Saturday","Sunday"))))  #creates variable wend (for weekend), where 1 = weekend, 0 = week. 
  dat <- mutate(dat, rushmod = rushhour)
  dat$rushmod[dat$wend==1] <- 3 
  dat <- mutate(dat,across(c(mem,wday,rushhour,rushmod), as.factor))
  
  return(dat)
}
```
```{r}
dat  <- preproc_dat(mydata)
 
```


### Data preparation and exploration
####  New variables
create the variables:  
  2 new variables for day and time, 
  rain( rain vs. no rain), 
  trip duration( trip_short = trip < 30min, trip_med = trip > 30min && trip < 60min), 
  month from day

```{r}
# 2 new variables for day and time
# Convert 'dep' column to date-time format
dat <- mutate(dat, dep = ymd_hms(dep))

# Create new columns for 'day' and 'time'
dat <- mutate(dat, day = as.Date(dep), time = format(dep, "%H:%M:%S"))

# Create a "rain vs no rain" variable
dat <- dat %>%
  mutate(rain = ifelse(prec > 0, "rain", "no rain"))

# Categorize trip duration into three categories: short, medium, and long
dat <- dat %>%
  mutate(trip_category = case_when(
    dur < 30 * 60 ~ "trip_short",       # Trip duration less than 30 minutes
    dur >= 30 * 60 & dur < 60 * 60 ~ "trip_med",   # Trip duration between 30 and 60 minutes
    dur >= 60 * 60 ~ "trip_long"        # Trip duration 60 minutes or more
  ))


# Extract the month from the "day" column
dat <- dat %>%
  mutate(month = month(day))  # Extract month and label with abbreviated month names

#factors 
dat$mem <- as.factor(dat$mem)
dat$trip_category <- as.factor(dat$trip_category)
dat$rain <- as.factor(dat$rain)


# View the first few rows of the dataset to confirm the new variables
head(dat)

```
```{r}
dat$trip_category <- as.factor(dat$trip_category)
levels(dat$trip_category)
```
##### Check the distribution of the rushmod variable
```{r}
# Check the distribution of the rushmod variable
table(dat$rushmod, dat$wend)

```


####Exploration

```{r}
summary(dat)
```


We have as much as member than non-member riders
```{r}
table(dat$mem)
```

```{r}
table(dat$wday)
```



### Graphs

```{r}
#Checking if any columsn have NA, nope! What a clean dataset. 
colSums(is.na(dat))

#Checking for data distribution
hist(dat$dur)
boxplot(dat$dur)
hist(log(dat$dur/60))         #seems like this normalizes the data nicely
hist(as.integer(dat$wday))    #all days appear equally represented
hist(dat$temp)                #temperature appears almost normally distributed
boxplot(dat$temp)
hist(dat$prec)                #Rain appears highly non normal as well as having multiple 0 entries, suggest to transform it into a factor of rain vs no rain. 
100*sum(dat$prec == 0)/length(dat$prec) #proportion or days without rain. 
```

## Data analysis
#### Statistic of trip duration for mem=1 and mem=0 

We need to calculate the average trip duration for both members and non-members. Then, we’ll adjust for weekend vs. non-weekend trips to see if the patterns hold

```{r echo=FALSE}
# Load necessary libraries
library(dplyr)

# Create a summary table for trip duration for members vs non-members
trip_duration_summary <- dat %>%
  group_by(mem) %>%
  summarize(
    count = n(),                          # Number of trips
    avg_duration = mean(dur, na.rm = TRUE), # Average trip duration
    median_duration = median(dur, na.rm = TRUE), # Median trip duration
    sd_duration = sd(dur, na.rm = TRUE),    # Standard deviation of trip duration
    min_duration = min(dur, na.rm = TRUE),  # Minimum trip duration
    max_duration = max(dur, na.rm = TRUE),  # Maximum trip duration
    Q1_duration = quantile(dur, 0.25, na.rm = TRUE),  # First quartile (25%)
    Q3_duration = quantile(dur, 0.75, na.rm = TRUE)   # Third quartile (75%)
  )

# View the summary table
print(trip_duration_summary)

```

#### Statistic on trip duration when we adjust for weekend vs. non-weekend usage (wend= 0 or 1)

```{r echo=FALSE}

# Summary of average trip duration by membership status
summary_mem_vs_nonmem <- dat %>%
  group_by(mem) %>%
  summarize(avg_duration = mean(dur, na.rm = TRUE))

# Summary of average trip duration by membership status and weekend
summary_mem_vs_nonmem_wend <- dat %>%
  group_by(mem, wend) %>%
  summarize(avg_duration = mean(dur, na.rm = TRUE))

# View summaries
print("Average trip duration for members vs non-members:")
print(summary_mem_vs_nonmem)

print("Average trip duration for members vs non-members, separated by weekend/non-weekend:")
print(summary_mem_vs_nonmem_wend)

```

Average trip duration for non members on weekend (wend=1) is higher than non members on weekdays (1153.80 vs 1090.89) . The same remarks goes for members, they tend to bike more on weekends. We also notice that non members bike more than members. Non members also bike for a longer time than members. Even after checking for weekends and weekdays. We can do a way ANOVA to assess the mean difference of duration for members vs non-members accounting for wend.

### Distribution of Duration with other variable

```{r}
library(ggplot2)
# library(tidyr)
ggplot(dat, aes(x = factor(mem, labels = c("Non-member", "Member")), fill = trip_category)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ wend) +  # Facet by weekend status (wend)
  labs(title = "Distribution of Trip Duration by Membership and Weekend Status",
       x = "Membership", y = "Count", fill = "Trip Duration Category") +
  scale_fill_manual(values = c("trip_short" = "#56B4E9", "trip_med" = "#009E73", "trip_long" = "#D55E00")) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



### Two-Sample t-test (Comparing Two Groups: Members vs. Non-members)
To assess if BIXI members have shorter trips than non-members

```{r}
# Perform a two-sample t-test

t_test_result <- t.test(dur ~ mem, data = dat)

# Print the result of the t-test
print(t_test_result)
```
Based on the very small p-value, we reject the null hypothesis and conclude that non-members take significantly longer trips on average compared to members. The difference in mean trip durations is statistically significant, with non-members taking trips that are 3.78 to 6.61 minutes longer on average than members. 
##### when include wend 

```{r}

# Model 1: Trip duration ~ membership (mem)
model1 <- lm(dur ~ mem, data = dat)

# Model 2: Trip duration ~ membership (mem) + weekend (wend) + interaction
model2 <- lm(dur ~ mem * wend, data = dat)

# Compare the two models using ANOVA
anova_comparison <- anova(model1, model2)

# Print the ANOVA comparison result
print(anova_comparison)

```
Non-significant p-value (0.5673): The addition of the weekend status (wend) and the interaction between membership and weekend status (mem * wend) does not significantly improve the model's ability to explain trip duration.
This means that, based on this analysis, weekend status (wend) does not have a significant effect on trip duration, nor does it significantly interact with membership (mem) to affect trip duration.


### With weather (rain or no rain)

the distribution of trip duration for members vs. non-members, broken down by whether the trip happened on a weekend or weekday and whether it was raining.


```{r}
ggplot(dat, aes(x = factor(mem, labels = c("Non-member", "Member")), fill = trip_category)) +
  geom_bar(position = "dodge") +
  facet_grid(rain ~ wend) +  # Switch the facets (rain on rows, wend on columns)
  labs(title = "Distribution of Trip Duration by Membership, Rain, and Weekend Status",
       x = "Membership", y = "Count", fill = "Trip Duration Category") +
  scale_fill_manual(values = c("trip_short" = "#56B4E9", "trip_med" = "#009E73", "trip_long" = "#D55E00")) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
# Model 3: Trip duration ~ membership (mem) + weekend (wend) + rain + interactions
model3 <- lm(dur ~ mem * wend * rain, data = dat)

# Perform the ANOVA to compare models
anova_comparison_rain <- anova(model2, model3)

# Print the ANOVA comparison result
print(anova_comparison_rain)


```
Non-significant p-value (0.6843): This means that adding rain and its interactions with membership and weekend status does not significantly improve the model's fit.
Therefore, rain does not have a statistically significant impact on trip duration in this dataset.
Based on this analysis, rain does not seem to be an important factor in explaining trip duration, and you can stick with the simpler Model 2 (which includes only membership and weekend status).

### With rushmod
The linear model will show if there are significant differences in trip durations between:

    AM rush hour (rushmod = 1).
    PM rush hour (rushmod = 2).
    Non-rush hour (rushmod = 3).
    
```{r}
# Filter the data to only include weekday trips (wend == 0)
weekday_data <- dat %>% filter(wend == 0)
# Convert rushmod to a factor for modeling
weekday_data$rushmod <- as.factor(weekday_data$rushmod)
# Fit a linear model to compare trip durations based on rush hour categories
rush_hour_model <- lm(dur ~ rushmod, data = weekday_data)

# Print the summary of the model
summary(rush_hour_model)

```
The p-values for both rushmod2 (PM rush hour) and rushmod3 (non-rush hour) are highly significant (p = 0.000814 and p = 0.000251, respectively). This suggests that both PM rush hour and non-rush hour trip durations are significantly longer than AM rush hour trip durations.
```{r}
# Perform Tukey HSD post-hoc test for pairwise comparisons
rushmod_comparison <- TukeyHSD(aov(rush_hour_model))

# Print the pairwise comparison results
print(rushmod_comparison)

```
PM rush vs. AM rush (2-1):

    Difference: 212.71 seconds longer in PM rush hour compared to AM rush hour.
    p-value: 0.0023 (significant), meaning the trip durations during PM rush hour are significantly longer than during AM rush hour.

Non-rush vs. AM rush (3-1):

    Difference: 232.72 seconds longer in non-rush hour compared to AM rush hour.
    p-value: 0.0007 (significant), meaning the trip durations during non-rush hours are significantly longer than during AM rush hour.

Non-rush vs. PM rush (3-2):

    Difference: 20.01 seconds longer in non-rush hour compared to PM rush hour.
    p-value: 0.9465 (not significant), meaning there is no significant difference between PM rush hour and non-rush hour trip durations.

Conclusion:

    AM rush hour trips are significantly shorter than both PM rush hour and non-rush hour trips.
    There is no significant difference between PM rush hour and non-rush hour trip durations.



